---
title: "Expectancy"
output: github_document
---
Data visualization in support of establishing cut scores for employment tests based on quantile assignment.  
[@AJThurston](twitter.com/AJThurston)

## Introduction

Expectancy charts are useful for establishing cut scores for selection (e.g., Stark et al. 2014, p. 160). This data visualization shows results by quantile groups, and are generally useful for communicating validity in terms of a correlation coefficient to non-technical audiences and have been found to be legally defensible in establishing cut scores for personnel decisions (Kehoe & Olson, 2005). This basic script will create quantiles based on the predicted criterion score and plot the mean actual criterion scores from each quantile. It includes the option to can export the quantile assignment back to a CSV file.

You should also note there are other ways to create expectancies and there are limitations to this method.  Historically, I-O psychologists have used Taylor-Russell tables to calculate these values, and there is a new algorithmic approach that allows for corrections for criterion unreliability (Cucina, Berger, & Busciglio 2017). 

## Setup and Libraries

```{r setup, message = FALSE, warning = FALSE}
library(summarytools)
library(formattable)
library(tidyverse)
library(ggplot2)
library(scales)
library(Cairo)
```

## Data Import

In this example, these simulated data have a correlation of r = .5 between predicted and actual criterion scores (N = 1000, M = 50, SD = 10). First, let's import the example data and quickly orient ourselves to it. There are three variables in the file:

1.  id: This is just an identifier or a unique participant ID
2.  actu: this is the acutal criterion score (e.g., actual job performance)
3.  pred: this is the predicted criterion score (e.g., predicted job perfromance from an employment test)

```{r data, message = FALSE, warning = FALSE}
# df <- read.csv("https://raw.githubusercontent.com/AJThurston/quantiles/master/expectancy.csv")
df <- read.csv("expectancy.csv")
head(df)

df %>%
  select(actu,pred) %>%
  descr(.)
```

## Parameters

Next, we'll need to set some parameters for the analysis and the graphics of the plot itself. The first parameter is simply how many quantiles we want to display, we could include only four (i.e., quartiles), this example uses five (i.e., quintiles), but I would recommend probably no more than 10 (i.e., deciles).

Additionally, you'll need to include the text size for the plot, the upper and lower limits for the y-axis, and the axis titles. For the limits, I typically recommend using the actual criterion mean minus two SD for your lower limit and two SD above the mean for the upper limit. Some could argue it should be 0 to 100 if that is the range of possible scores. I would argue my limits cover over 95% of the possible scores. Whatever limits you choose, ensure you have some justification.

```{r parameters, message = FALSE, warning = FALSE}
quants  <-  5                              #Number of quantiles
txt.siz <- 12                              #Size for all text in the plot
y.ll    <- 30                              #Y-axis lower limit          
y.ul    <- 70                              #Y-axis upper limit
x.title <- "Predicted Criterion Quantiles" #X-axis title
y.title <- "Mean Actual Criterion Score"   #Y-axis title
```

## Calculating Quantiles

Based on the number of quantiles indicated in the parameter above, now we actually need to calculate the thresholds between each of the quantiles then assign each predicted score to a quantile group. The first bit of code here calculates the quantiles with the `quantile` function, then the code below takes the predicted values, compares them to the quantiles, and stores the quantile assignment as a new vairable using the `cut` function. Finally, the average actual criterion score for each quantile group is shown.

```{r quantiles, message = FALSE, warning = FALSE}
quantiles <- quantile(df$pred, probs = seq(0,1,1/quants))
quantiles

df$quant <- df$pred %>%
  cut(., breaks = quantiles, include.lowest=TRUE) %>%
  as.numeric()

head(df)

df %>%
  group_by(quant) %>%
  summarize(m = mean(actu))
```

## Expectancy Plot

In many cases it is easier to simply take these values and display them in, for example, a PowerPoint plot. However, if you have to automate this tast for multiple criteria, the code below may be useful for this purpose.

```{r expectancy, message = FALSE, warning = FALSE}
p <- ggplot(df)
p <- p + scale_y_continuous(name=y.title, limits = c(y.ll,y.ul), oob = rescale_none)
p <- p + scale_x_continuous(name=x.title, oob = rescale_none)
p <- p + geom_bar(aes(x = quant, y = actu), 
                  position = "dodge", 
                  stat = "summary", 
                  fun = "mean",
                  fill = '#336666',
                  width = .5)
p <- p + geom_text(aes(x = quant, y = actu, label = paste0(round(..y..,0),"%")), 
                   stat = "summary", 
                   fun = "mean",
                   vjust = -1)
p <- p + theme(text = element_text(size = txt.siz),
               panel.background = element_rect(fill = "white", color = "black"),
               panel.grid = element_blank(),
               axis.text.y = element_text(color = 'black'),
               axis.text.x = element_text(color = 'black'))
p
```

## Export

These are some options for exporting your expectancy chart and the data for use in other software programs.

```{r export, message=FALSE, warning=FALSE}
ggsave("expectancy.png", 
       plot = p, 
       scale = 1, 
       width = 6.5, 
       height = 4, 
       units = "in",
       dpi = 300,
       type = "cairo-png")

write.csv(data, "expectancy_appended.csv")
```

If you would prefer the algorithmic approach, I have copied the code from Cucina et al.(2017) for ease of use here: https://github.com/AJThurston/expectancy/miwa_expectancy.R

## References
Cucina, J., Berger, J., & Busciglio, H. (2017). Communicating Criterion-Related Validity Using Expectancy Charts: A New Approach. Personnel Assessment and Decisions, 3(1). https://doi.org/10.25035/pad.2017.001


Kehoe, J. F., & Olson, A. (2005). Cut scores and employment discrimination litigation. In F. J. Landy (Ed.), *Employment discrimination litigation: Behavioral, quantitative, and legal perspectives* (pp. 410–449). San Francisco, CA: Jossey-Bass.

Stark, S., Chernyshenko, O. S., Drasgow, F., Nye, C. D., White, L. A., Heffner, T., & Farmer, W. L. (2014). From ABLE to TAPAS: A New Generation of Personality Tests to Support Military Selection and Classification Decisions. Military Psychology, 26(3), 153–164. https://doi.org/10.1037/mil0000044

